---
title: "3220 Project BACKUP"
author: "Maximilian Van Zandt"
date: "2023-11-25"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

### **Data Cleaning**

```{r, echo=FALSE, warning=FALSE,results=FALSE}
library(dplyr)
library(tidyverse)
library(tinytex)
library(RColorBrewer)
library(corrplot)
library(Hmisc)
library(olsrr)
library(car)
library(MASS)
Data <- read.csv("data/processed/merged_data.csv")
```


```{r, echo=FALSE, warning=FALSE,results=FALSE}
library(ggplot2)

ggplot(Data, aes(x = DeathRate)) +
  geom_histogram(binwidth=100, fill = "lightblue", color = "black") +
  labs(x = "Death Rate", y = "Frequency") +
  ggtitle("Histogram of Death Rate")

#Correlation matrix between medicare enrollment rates and death rate
CorMD<-cor(Data$MedicareEnr, Data$DeathRate)
cat("Correlation between Medicare Enrollment Rate and Death Rate:", CorMD)


#Correlation between drug overdose rate and death rate
CorDOD<-cor(Data$OverdoseDeaths, Data$DeathRate)
cat("Correlation between Drug Overdose Rate and Death Rate:", CorDOD)

#Side-by-side boxplot for COVID and death rate (Research Question - Numeric)
ggplot(Data, aes(x = LiftCOVID, y = DeathRate, fill = LiftCOVID)) +
  geom_boxplot(color = "Black",lwd=0.5)+labs(x = "Lifted COVID Restrictions", y = "Death Rate", title = "Lifted COVID Restrictions on Death Rate", fill="Legend")+theme(plot.title = element_text(hjust = 0.5))

#Scatter plot between Medicaid spending per capita and death rate
ggplot(data=Data, aes(x=MedicaidSpend, y=DeathRate))+labs(x="Medicaid Spending per Capita",y="Death Rate",title="Medicaid Spending per Capita on Death Rate")+geom_point()

#Interaction plot between political majority and COVID restrictions on death rate
ggplot(Data, aes(x = LiftCOVID, color = PoliticalMaj, group = PoliticalMaj, y = DeathRate))+stat_summary(fun.y = mean, geom = "point")+stat_summary(fun.y = mean, geom = "line")

#Side-by-side boxplot between the percentage of the senior population and the death rate
ggplot(Data, aes(x = SeniorPop, y = DeathRate, fill = SeniorPop)) +
  geom_boxplot(color = "Black",lwd=0.5)+labs(x = "Percentage of Senior Population", y = "Death Rate", title = "Percentage of Senior Population on Death Rate", fill="Legend")+theme(plot.title = element_text(hjust = 0.5))


#Correlation between obesity prevalence and death rate
CorOPD<-cor(Data$ObesityPrev, Data$DeathRate)
cat("Correlation between Obesity Prevalence and Death Rate:", CorOPD)


#Correlation between population density and death rate
CorPD<-cor(Data$PopPerCap, Data$DeathRate)
cat("Correlation between Population Density and Death Rate:", CorPD)


```

```{r, echo=FALSE, warning=FALSE,results=FALSE}
names(Data)
```

### **Part 2: Analysis and Model Building**

### **Stage 1: Quantitative Variables**

*Initial:* $DeathRate = \beta_0+\beta_1ObesityPrev+\beta_2OverdoseDeaths+\beta_3MedicareEnr+\beta_4MedicaidSpend+\beta_5PopPerCap$

**EDA: Scatterplots and Multicolinearity**

```{r}
#Scatter plots for quantitative variables

for (i in names(Data)[c(3:5, 10:11)]) {
  print(ggplot(data = Data, aes_string(x = i, y = "DeathRate")) +
    labs(x = i, y = "Death Rate") +
    geom_point())
}

#Multicolinearity
Datacor<-round(cor(Data[,-c(1,2,6,7,8,9)]),4)
Datacor
corrplot(Datacor)
#VIF
Datamod1<-lm(DeathRate~ObesityPrev+OverdoseDeaths+MedicareEnr+MedicaidSpend+PopPerCap,data=Data)
vif(Datamod1)
mean(vif(Datamod1))


#Correlations for quantitative variables
round(cor(Data[c(3:5,10:11)],Data$DeathRate),3)
```

Multicolinearity Conclusion: No evidence of severe multicolinearity. No individual VIFs greater than 10 and average VIF is less than 3.

**Variable Screening: Stepwise Regression**

```{r}
#Variable Screening (Stepwise Regression)
ols_step_both_p(Datamod1,pent=0.1,prem=0.3,details=F)
```

Stepwise Resulting Model: $DeathRate = \beta_0+\beta_1ObesityPrev+\beta_2OverdoseDeaths+\beta_5PopPerCap$


**Adding Higher Order Term (PopPerCap)**

```{r}
Datamod2<-lm(DeathRate~ObesityPrev+OverdoseDeaths+PopPerCap+I(PopPerCap^2),data=Data)       
summary(Datamod2)
```

Significant so ADD (t test)

**Adding Quant x Quant Interaction**

```{r}
Datamod3<-lm(DeathRate~ObesityPrev+OverdoseDeaths+PopPerCap+ObesityPrev*OverdoseDeaths,data=Data)       
summary(Datamod3)
```

Not Significant so REMOVE (t test)

**Final Quant Model**

```{r}
Datamod2<-lm(DeathRate~ObesityPrev+OverdoseDeaths+PopPerCap+I(PopPerCap^2),data=Data)       
summary(Datamod2)
```

*Global F Test:*

Hypothesis: Null - model is not adequate Alternative - model is adequate

Decision: 0.0001\<0.05 -\> REJECT H0

Conclusion: The model with obesity prevalence, overdose death rate, population per capita, and population per capita squared is adequate at predicting death rate.

**Quantitative Variables Summary:**

Initial: $DeathRate = \beta_0+\beta_1ObesityPrev+\beta_2OverdoseDeaths+\beta_3MedicareEnr+\beta_4MedicaidSpend+\beta_5PopPerCap$

Result from Stepwise: $DeathRate = \beta_0+\beta_1ObesityPrev+\beta_2OverdoseDeaths+\beta_3PopPerCap$

Test higher order term: $\beta_4PopPerCap^2$ -- Significant (individual t test)

Test Quant x Quant Interaction: $\beta_5ObesityPrev*OverdoseDeaths$ -- NOT Significant (individual t test)

Final: $DeathRate = \beta_0+\beta_1ObesityPrev+\beta_2OverdoseDeaths+\beta_3PopPerCap+\beta_4PopPerCap^2$

### **Stage 2: Qualitative Variables**

*Initial:* $DeathRate = \beta_0+\beta_1ObesityPrev+\beta_2OverdoseDeaths+\beta_3PopPerCap+\beta_4PopPerCap^2+\beta_5SeniorPopM+\beta_6SeniorPopH+\beta_6PoliticalMajR+\beta_7LiftCOVIDY$

where SeniorPopM = 1 if Medium, 0 otherwise

SeniorPopH = 1 if High, 0 otherwise

PoliticalMajR = 1 if Republican, 0 otherwise

LiftCOVIDY = 1 if Yes, 0 otherwise

**EDA: Side-by-side Boxplots**

```{r}
#Boxplots for Qualitative Variables
cbp1 <- c("#E69F00", "#56B4E9", "#009E73",
          "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

Data$SeniorPop <- factor(Data$SeniorPop, levels = c("Low", "Medium", "High"))

ggplot(Data, aes(x = SeniorPop, y = DeathRate, fill = SeniorPop)) +
  geom_boxplot(color = "Black",lwd=0.5)+labs(x = "Percentage of Senior Population", y = "Death Rate", title = "Percentage of Senior Population on Death Rate", fill="Legend")+theme(plot.title = element_text(hjust = 0.5))+scale_fill_manual(values = cbp1)

ggplot(Data, aes(x = LiftCOVID, y = DeathRate, fill = LiftCOVID)) +
  geom_boxplot(color = "Black",lwd=0.5)+labs(x = "Lifted COVID Restrictions", y = "Death Rate", title = "Lifted COVID Restrictions on Death Rate", fill="Legend")+theme(plot.title = element_text(hjust = 0.5))+scale_fill_manual(values = cbp1)

ggplot(Data, aes(x = PoliticalMaj, y = DeathRate, fill = PoliticalMaj)) +
  geom_boxplot(color = "Black",lwd=0.5)+labs(x = "Political Majority", y = "Death Rate", title = "Political Majority on Death Rate", fill="Legend")+theme(plot.title = element_text(hjust = 0.5))+scale_fill_manual(values = c("D" = "#56B4E9", "R" = "#FF6666"))
```

```{r}
Datamod4<-lm(DeathRate~ObesityPrev+OverdoseDeaths+PopPerCap+I(PopPerCap^2)+SeniorPop+PoliticalMaj+LiftCOVID,data=Data)       
summary(Datamod4)
#Nested F Test
NestDatamod<-lm(DeathRate~ObesityPrev+OverdoseDeaths+PopPerCap+I(PopPerCap^2)+PoliticalMaj+LiftCOVID,data=Data) 
anova(NestDatamod,Datamod4)
```

Initial: $DeathRate = \beta_0+\beta_1ObesityPrev+\beta_2OverdoseDeaths+\beta_3PopPerCap+\beta_4PopPerCap^2+\beta_5SeniorPopM+\beta_6SeniorPopH+\beta_6PoliticalMajR+\beta_7LiftCOVIDY$

Final: $DeathRate = \beta_0+\beta_1ObesityPrev+\beta_2OverdoseDeaths+\beta_3PopPerCap+\beta_4PopPerCap^2$

Note: NO qualitative variables were signficant.

Datamod5<-lm(DeathRate~ObesityPrev+OverdoseDeaths+PopPerCap+I(PopPerCap^2),data=Data)       


### **Stage 3: Interactions**

```{r}
#Interactions
ggplot(Data, aes(x = MedicaidSpend, y = DeathRate, color = PoliticalMaj))+geom_point()+labs(x="Medicaid Spending", y="Death Rate")+scale_color_manual(values = c("D" = "#56B4E9", "R" = "#FF6666"))+theme(legend.position = "top")

ggplot(Data, aes(x = MedicareEnr, y = DeathRate, color = PoliticalMaj))+geom_point()+labs(x="Medicare Enrollment", y="Death Rate")+scale_color_manual(values = c("D" = "#56B4E9", "R" = "#FF6666"))+theme(legend.position = "top")

ggplot(Data, aes(x = MedicareEnr, y = DeathRate, color = SeniorPop))+geom_point()+labs(x="Medicare Enrollment", y="Death Rate")+scale_fill_manual(values = cbp1)+theme(legend.position = "top")

ggplot(Data, aes(x = MedicareEnr, y = DeathRate, color = PoliticalMaj))+geom_point()+labs(x="Medicare Enrollment", y="Death Rate")+scale_color_manual(values = c("D" = "#56B4E9", "R" = "#FF6666"))+theme(legend.position = "top")
```

**Conclusion:** No Qual x Quant interactions because all of the qualitative variables have been removed due to lack of significance. Additionally, the EDA does not provide evidence to indicate an interaction.

Potential Title: What Kills? Factors that Affect State Death Rates

### **Regression Assumptions**
```{r}
Datamod5<-lm(DeathRate~ObesityPrev+OverdoseDeaths+PopPerCap+I(PopPerCap^2),data=Data)
summary(Datamod5)
#Lack of Fit (Residual Plots)
residualPlots(Datamod5,tests=F)

#Constant Variance (Residual Plots and Residual vs Fitted)
residualPlots(Datamod5,tests=F)
plot(Datamod5, which=1)

#Normality (QQ Plot and Histogram of Residuals)
plot(Datamod5, which=2)
hist(residuals(Datamod5))
```
Lack of Fit (Residual Plots): This assumption does not appear to be violated because there doesn't appear to be any trends that can be modeled by a transformation (quadratic, logarithmic, etc.).

Constant Variance (Residual Plots and Residual vs Fitted): Many of the residual plots including residual vs fitted, residual vs horsepower, residual vs. compression ratio show a fanning out pattern and the values for the residual vs. compression ratio plot are clustered on either end and inconsistently. This indicates a violation of the constant variance assumption.

Normality (QQ Plot and Histogram of Residuals): The Q-Q plot shows the residuals deviate from the line in both directions. Additionally, the histogram of the residuals is close to being normally distributed, but has outliers that force a skew. In conclusion the assumption is violated, but might not be of serious concern.

**Conclusion:** Transform the PopPerCap variable. We'll use the log transformation.

**Transformation and Re-test Assumptions**
```{r}
Datamod6<-lm(DeathRate~ObesityPrev+OverdoseDeaths+I(log(PopPerCap)),data=Data)
summary(Datamod6)

#Lack of Fit (Residual Plots)
residualPlots(Datamod6,tests=F)

#Constant Variance (Residual Plots and Residual vs Fitted)
residualPlots(Datamod6,tests=F)
plot(Datamod6, which=1)

#Normality (QQ Plot and Histogram of Residuals)
plot(Datamod6, which=2)
hist(residuals(Datamod6))
```

The constant variance assumption has been met now. Therefore, the log transformation successfully fixed the model and we can conclude that it is the final model of our analysis.

**FINAL MODEL:** $DeathRate = \beta_0+\beta_1ObesityPrev+\beta_2OverdoseDeaths+\beta_3log(PopPerCap)$

### **Specifics for additional technique**

1.  Variable Screening (Stepwise Regression) - Done in previous section

2.  Influential Observations

```{r}
Datamod6<-lm(DeathRate~ObesityPrev+OverdoseDeaths+I(log(PopPerCap)),data=Data)  
plot(Datamod6,which=4)
influencePlot(Datamod6,fill=F)
```


3.  Weighted Least Squares (Weighting each observation)

```{r}
wt<-1/lm(abs(Datamod6$residuals)~Datamod6$fitted.values)$fitted.values^2

#perform weighted least squares regression
Datamod6 <- lm(DeathRate~ObesityPrev+OverdoseDeaths+I(log(PopPerCap)),data=Data, weights=wt)
summary(Datamod6)

#Lack of Fit (Residual Plots)
residualPlots(Datamod6,tests=F)

#Constant Variance (Residual Plots and Residual vs Fitted)
residualPlots(Datamod6,tests=F)
plot(Datamod6, which=1)

#Normality (QQ Plot and Histogram of Residuals)
plot(Datamod6, which=2)
hist(residuals(Datamod6))
```
